---
title: "Working with dataframes in R"
author: "Vicki M. Zhang"
urlcolor: blue
output:
  pdf_document: default
  word_document: default
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

## Getting ready to code

Note that today's lesson is largely adapted from
[EEB313](https://uoftcoders.github.io/rcourse/lec04-dplyr.html), taught
in 2021 by a group of awesome grad students!

We'll be using the animal survey dataset.

```{r}
# URL of the dataset on Figshare
file_url <- "https://ndownloader.figshare.com/files/2292169"

# Destination file path on your local machine
destination_file <- "portal_data_joined.csv"

# Download the file
download.file(file_url, destination_file)

# Load the data into a data frame (using readr package for read_csv)
library(readr)
surveys <- read_csv(destination_file)
```

## Working with a subset of the data

We'll start by doing our checks.

```{r}
library(tidyverse)
str(surveys)
head(surveys)
```

## Data wrangling with dplyr

Wrangling here is used in the sense of maneuvering, managing,
controlling, and turning your data upside down and inside out to look at
it from different angles in order to understand it. The package `dplyr`
provides easy tools for the most common data manipulation tasks. It is
built to work directly with data frames, with many common tasks
optimized by being written in a compiled language (C++), this means that
many operations run much faster than similar tools in R. An additional
feature is the ability to work directly with data stored in an external
database, such as SQL-databases. The ability to work with databases is
great because you are able to work with much bigger datasets (100s of
GB) than your computer could normally handle. We will not talk in detail
about this in class, but there are great resources online to learn more
(e.g. [this lecture from Data
Carpentry](https://datacarpentry.org/R-ecology-lesson/05-r-and-databases.html)).

### Selecting columns and filtering rows

We're going to learn some of the most common `dplyr` functions:
`select()`, `filter()`, `mutate()`, `group_by()`, and `summarise()`. To
select columns of a data frame, use `select()`. The first argument to
this function is the data frame (`surveys`), and the subsequent
arguments are the columns to keep.

```{r}
select(surveys, plot_id, species_id, weight, year)
```

*Note: Unlike in base R, here we don't need to use quotations around
column names.*

To choose rows based on a specific criteria, use `filter()`:

```{r}
filter(surveys, year == 1995)
```

*Note2: To check for equality, R requires **two** equal signs (`==`).
This is different than object assignment where we use ('\<-') or ('=')
to assign values to an object. With `filter` we want to pull out all
rows where `year` is equal to 1995 not assign the value 1995 to an
object named `year`, so we use the `==` symbol.*

*Note3: In general, when you want to find all rows that equal a numeric
number you don't have to use quotes. However, when you want to find all
rows that equal a character you do need to put quotes around the value
(for instance `taxa == "Rodent"`).*

#### An aside on conditionals

Within `filter` you might want to filter rows using conditionals. Basic
conditionals in R are broadly similar to how they're expressed
mathematically:

```{r}
2 < 3
5 > 9
5 == 5
```

However, there are a few idiosyncrasies to be mindful of for other
conditionals:

```{r}
2 != 3 # Not equal
2 <= 3 # Less than or equal to
5 >= 9 # Greater than or equal to
```

Finally, the `%in%` operator is used to check for membership:

```{r}
2 %in% c(2, 3, 4) # Checks whether 2 is in c(2, 3, 4), returns logical vector
```

All of the above conditionals are compatible with `filter`, with the key
difference being that `filter` expects column names as part of
conditional statements instead of individual numbers.

### Chaining functions together using pipes

But what if you wanted to select and filter at the same time? There are
three ways to do this: use intermediate steps, nesting functions, or
pipes.

With intermediate steps, you essentially create a temporary data frame
and use that as input to the next function:

```{r}
temp_df <- select(surveys, plot_id, species_id, weight, year)
filter(temp_df, year == 1995)
```

This can quickly clutter up your workspace with lots of objects...

You can also nest functions (i.e., one function inside of another).

```{r}
filter(select(surveys, plot_id, species_id, weight, year), year == 1995)
```

This is handy, but can be difficult to read if too many functions are
nested as they are evaluated from the inside out.

The last option, *pipes*, are a fairly recent addition to R. Pipes let
you take the output of one function and send it directly to the next,
which is useful when you need to do many things to the same dataset.
Pipes in R look like `%>%` and are made available via the `magrittr`
package that is a part of the `tidyverse`. If you use RStudio, you can
type the pipe with <kbd>Ctrl/Cmd</kbd> + <kbd>Shift</kbd> +
<kbd>M</kbd>.

```{r}
surveys %>% 
    select(., plot_id, species_id, weight, year) %>% 
    filter(., year == 1995)
```

The `.` refers to the object that is passed from the previous line. In
this example, the data frame `surveys` is passed to the `.` in the
`select()` statement. Then, the modified data frame (which is the result
of the `select()` operation) is passed to the `.` in the filter()
statement. Put more simply: whatever was the result from the line above
will be used in the current line.

Since it gets a bit tedious to write out all the dots, `dplyr` allows
for them to be omitted. In the `dplyr` family of functions, the first
argument is always a data frame, and by default the pipe will pass the
output from the line above to this argument. The chunk below, with the
`.` omitted, gives the same output as the one above:

```{r}
surveys %>% 
    select(plot_id, species_id, weight, year) %>% 
    filter(year == 1995)
```

If this runs off your screen and you just want to see the first few
rows, you can use a pipe to view the `head()` of the data. Pipes work
with non-`dplyr` functions, too, as long as either the `dplyr` or
`magrittr` package is loaded.

```{r}
surveys %>% 
    select(plot_id, species_id, weight, year) %>% 
    filter(year == 1995) %>%
    head()
```

If we wanted to create a new object with this smaller version of the
data, we could do so by assigning it a new name:

```{r}
surveys_1995 <- surveys %>% 
    select(plot_id, species_id, weight, year) %>% 
    filter(year == 1995)

surveys_1995
```

#### Challenge

Use the pipe to subset the data frame, keeping only rows where `weight`
is less than 10, and only the columns `species_id`, `sex`, and `weight`.

```{r, include=FALSE}
surveys %>%
  filter(weight < 10) %>%
  select(species_id, sex, weight)
```

We could write a single expression to filter for several criteria,
either matching *all* criteria (`&`) or *any* criteria (`|`):

```{r}
surveys %>% 
    filter(taxa == 'Rodent' & sex == 'F') %>% 
    select(sex, taxa)
```

```{r}
surveys %>% 
    filter(species == 'clarki' | species == 'audubonii') %>% 
    select(species, taxa)
```

### Creating new columns with mutate

Frequently, you'll want to create new columns based on the values in
existing columns. For instance, you might want to do unit conversions,
or find the ratio of values in two columns. For this we'll use
`mutate()`.

To create a new column of weight in kg:

```{r}
surveys %>%
    mutate(weight_kg = weight / 1000)
```

You can also create a second new column based on the first new column
within the same call of `mutate()`:

```{r}
surveys %>%
    mutate(weight_kg = weight / 1000,
           weight_kg2 = weight_kg * 2)
```

We can see that there is some `NA`s in our new column. If we wanted to
remove those we could insert a `filter()` in the chain, paired with the
`!is.na` notation we learned in the last lecture:

```{r}
surveys %>%
    filter(!is.na(weight)) %>%
    mutate(weight_kg = weight / 1000)
```

#### Challenge

Create a new data frame from the `surveys` data that meets the following
criteria: contains only the `species_id` column and a new column called
`hindfoot_half` containing values that are half the `hindfoot_length`
values. In this `hindfoot_half` column, there should be no `NA`s and all
values should be less than 30. (Hint: think about how the commands
should be ordered to produce this data frame.)

```{r, include=FALSE}
## Answer
surveys_hindfoot_half <- surveys %>%
    filter(!is.na(hindfoot_length)) %>%
    mutate(hindfoot_half = hindfoot_length / 2) %>%
    filter(hindfoot_half < 30) %>%
    select(species_id, hindfoot_half)
```

## Split-apply-combine techniques in dplyr

Many data analysis tasks can be approached using the
*split-apply-combine* paradigm: split the data into groups, apply some
analysis to each group, and then combine the results.

`dplyr` facilitates this workflow through the use of `group_by()` and
`summarize()`, which collapses each group into a single-row summary of
that group. The arguments to `group_by()` are the column names that
contain the *categorical* variables for which you want to calculate the
summary statistics. Let's view the mean `weight` by sex.

```{r}
surveys %>%
    group_by(sex) %>%
    summarize(mean_weight = mean(weight))
```

The mean weights become `NA` since there are individual observations
that are `NA`... Oh the NAs! Let's remove those observations.

```{r}
surveys %>%
    filter(!is.na(weight)) %>%
    group_by(sex) %>%
    summarize(mean_weight = mean(weight))
```

There is one row here that is neither male nor female. These are
observations where the animal escaped before the sex could be
determined. Let's remove those as well.

```{r}
surveys %>%
    filter(!is.na(weight) & !is.na(sex)) %>%
    group_by(sex) %>%
    summarize(mean_weight = mean(weight))
```

You can also group by multiple columns:

```{r}
surveys %>%
    filter(!is.na(weight) & !is.na(sex)) %>%
    group_by(genus, sex) %>%
    summarize(mean_weight = mean(weight))
```

Since we will use the same filtered and grouped data frame in multiple
code chunks below, we could assign this subset of the data to a new name
and use this data frame in the subsequent code chunks instead of typing
out the functions each time.

```{r}
filtered_surveys <- surveys %>%
    filter(!is.na(weight) & !is.na(sex)) %>%
    group_by(genus, sex)
```

Once the data are grouped, you can also summarize multiple variables at
the same time. For instance, we could add a column indicating the
minimum weight for each species for each sex:

```{r}
filtered_surveys %>%
    summarize(mean_weight = mean(weight),
              min_weight = min(weight))
```

#### Challenge

1.  Use `group_by()` and `summarize()` to find the mean hindfoot length
    for the species *Ammospermophilus harrisi*.

2.  What was the heaviest animal measured in 1979?

```{r, include=FALSE}
# Answer 1
surveys %>%
    filter(!is.na(hindfoot_length)) %>%
    group_by(species) %>%
    summarize(mean_hindfoot_length = mean(hindfoot_length)) # 31

# Answer 2
surveys %>%
    filter(!is.na(weight)) %>%
    filter(year == 1979) %>% 
    filter(weight == max(weight)) %>% 
    select(genus, species) # Neotoma albigula (white-throated woodrat)
```

### Using tally to summarize categorical data

When working with data, it is also common to want to know the number of
observations found for each factor or combination of factors. For this,
`dplyr` provides `tally()`. For example, if we want to group by taxa and
find the number of observations for each taxa, we would do:

```{r}
surveys %>%
    group_by(taxa) %>%
    tally()
```

We can also use `tally()` when grouping on multiple variables:

```{r}
surveys %>%
    group_by(taxa, sex) %>%
    tally()
```

Here, `tally()` is the action applied to the groups created by
`group_by()` and counts the total number of records for each category.

If there are many groups, `tally()` is not that useful on its own. For
example, when we want to view the five most abundant species among the
observations:

```{r}
surveys %>%
    group_by(species) %>%
    tally()
```

Since there are 32 rows in this output, we would like to order the table
to display the most abundant species first. In `dplyr`, we say that we
want to `arrange()` the data.

```{r}
surveys %>%
    group_by(species) %>%
    tally() %>%
    arrange(n) # `n` is the name of the column `tally` generated
```

Still not that useful. Since we are interested in the most abundant
species, we want to display those with the highest count first, in other
words, we want to arrange the column `n` in descending order:

```{r}
surveys %>%
    group_by(species) %>%
    tally() %>%
    arrange(desc(n)) %>%
    head(5)
```

If we want to include more attributes (like `taxa` and `genus`) about
these species, we can include these in the call to `group_by()`:

```{r}
surveys %>%
    group_by(species, taxa, genus) %>%
    tally() %>%
    arrange(desc(n)) %>%
    head(5)
```

Be careful not to include anything that would split the group into
subgroups, such as `sex`, `year` etc.

#### Challenge

How many individuals were caught in the rodent enclosure plot type?

```{r, include=FALSE}
surveys %>%
    group_by(plot_type) %>%
    tally()
```

#### Take-home challenge

You saw above how to count the number of individuals of each `sex` using
a combination of `group_by()` and `tally()`. How could you get the same
result using `group_by()` and `summarize()`? (Hint: see `?n`.)

```{r, include=FALSE}
surveys %>%
  group_by(sex) %>%
  summarize(n = n())
```

## Reshaping with pivot_longer and pivot_wider

The survey data presented here is almost in what we call a *long* format
-- every observation of every individual is its own row. This is an
ideal format for data with a rich set of information per observation. It
makes it difficult, however, to look at the relationships between
measurements across plots. For example, what is the relationship between
mean weights of different genera across all plots?

To answer that question, we want each plot to have a single row, with
all of the measurements in a single plot having their own column. This
is called a *wide* data format. For the `surveys` data as we have it
right now, this is going to be one heck of a wide data frame! However,
if we were to summarize data within plots and species, we can reduce the
dataset and begin to look for some relationships we'd want to examine.

We will be using the functions `pivot_wider` and `pivot_longer`.

Let's start by using `dplyr` to create a data frame with the mean body
weight of each genus by plot.

```{r}
surveys_gw <- surveys %>%
    filter(!is.na(weight)) %>%
    group_by(genus, plot_id) %>%
    summarize(mean_weight = mean(weight))

head(surveys_gw)
```

### Long to Wide with `pivot_wider`

Now, to make this long data wide, we use `spread` from `tidyr` to spread
out the different taxa into columns. `spread` takes three arguments: the
data, the *key* column (or column with identifying information), and the
*values* column (the one with the numbers/values). We'll use a pipe so
we can ignore the data argument.

```{r}
surveys_gw_wide <- surveys_gw %>%
  pivot_wider(names_from = genus, values_from = mean_weight) 

head(surveys_gw_wide)
```

Notice that some genera have `NA` values. That's because some of those
genera don't have any record in that plot. Sometimes it is fine to leave
those as `NA`. Sometimes we want to fill them as zeros, in which case we
would add the argument `fill=0`.

```{r}
surveys_gw %>%
  pivot_wider(names_from = genus, values_from = mean_weight,
              values_fill = 0) %>% 
  head()
```

Now we can go back to our original question: what is the relationship
between mean weights of different genera across all plots? We can easily
see the weights for each genus in each plot!

### Wide to long with `pivot_longer`

What if we had the opposite problem, and wanted to go from a wide to
long format? For that, we use `pivot_longer` to sweep up a set of
columns into one key-value pair. We give it the arguments of a new key
and value column name, and then we specify which columns we either want
or do not want gathered up. So, to go backwards from `surveys_gw_wide`,
and exclude `plot_id` from the gathering, we would do the following:

```{r}
surveys_gw_long <- surveys_gw_wide %>%
  pivot_longer(names_to = "genus", values_to = "mean_weight", cols = -plot_id) 

head(surveys_gw_long)
```

Note that now the `NA` genera are included in the long format. Going
from wide to long to wide can be a useful way to *balance* out a dataset
so every replicate has the same composition.

We could also have used a specification for what columns to include.
This can be useful if you have a large number of identifying columns,
and it's easier to specify what to gather than what to leave alone. And
if the columns are sequential, we don't even need to list them all out -
just use the `:` operator! Say we only wanted the columns from `Baiomys`
to `Onychomys` to be gathered together.

```{r}
surveys_gw_wide %>%
  pivot_longer(names_to = "genus", values_to = "mean_weight", cols = Baiomys:Onychomys) %>% 
  head()
```

#### Challenge

Starting with the `surveys_gw_wide` dataset, how would you create a new
dataset that gathers the mean weight of all the genera (excluding NAs)
except for the genus `Perognathus`?

```{r, include=FALSE}
surveys_gw_wide %>%
  pivot_longer(names_to = "genus", values_to = "mean_weight",
              cols = c(-plot_id, -Perognathus)) %>%
  filter(!is.na(mean_weight)) %>%
  head()
```
